Explainable AI with Probabilistic Graphical Models

### 2019-11-13 literature-review/explaining-mpe.tex
Thesis complete

% !TEX root = thesis-thomas-tiotto.tex

\section{Explaining the Most Probable Explanation}\label{sec:explaining-mpe}
\cite{Butz2018}

>>>>

\section{\enquote{Explaining the Most Probable Explanation}} \label{sec:explaining-the-most-probable-explanation}
The paper \enquote{Explaining the Most Probable Explanation} by \citet{Butz2018} places itself in the literature concerned with the explainability of Bayesian networks.
In particular, taking the classification proposed by \citet{lacave2002review} presented in Section \ref{sec:explainability-in-bayesian-networks}, it attempts to define a \textit{linguistic explanation} of the \textit{evidence} and of the \textit{reasoning}.
It differs from the previous attempt to define the explanation of the \textit{evidence} given by \citet{lacave2002review} and in other works, in that the paper is not concerned with finding the most probable assignment of variables that would explain the given evidence but, rather, the inverse problem.
By starting with the evidence and finding a maximally probable configuration, the authors hope \enquote{to look at the complete scenario to get an overview before deciding which variables should be focused on}; i.e., the goal appears to be to give the user an overview of the situation.

The initial claim of the paper is that BNs are still difficult to interpret for domain experts, even though these models provide a graphical structure to the \textit{knowledge base}.
The examples brought to justify the claim are that edges in the graph do not necessarily represent causal dependencies and that d-separation (Definition \ref{def:d-separation}) may be confusing.
The authors plan to address this claim by constructing a \textit{dialogue} with the user and thus to continue in the long tradition of dialogical approaches to explaining BNs, many of which are presented in \citep{lacave2002review}.

The defining characteristic of their approach is that the domain expert is able to \enquote{argue} with the MPE and investigate alternative explanations.
The complete methodology, executed over three steps, is shown in Figure \ref{fig:butz-methodology}.
The first step is the construction of the \enquote{knowledge base}, which is nothing else than a probability tree representing a \enquote{chain of deduction} constructed following the strongest probabilistic dependencies between variables in the BN.
Such a \textit{knowledge base} is convenient because the document plan for the Natural Language Generation step is directly derived from it.
One issue that is immediately apparent is that this greedy approach does not \enquote{generate the MPE solution} as the authors claim.
This does not discredit the argumentative method as a whole, as \textit{it is not necessary for the user to be arguing the MPE to derive a good explanation}; this ties into one of the main findings in the previous sections that many xAI researchers are only focusing on one half of the explanation.
A good explanation is not given only by its formal properties but, most importantly, by how well it acts as an interface between the real \textit{user} and the model.
This is what \citet{abdul2018trends} mean when they lament that \enquote{despite their mathematical rigour, these works \textit{[referring to the existing explainability methods]} suffer from a lack of usability, practical interpretability and efficacy on real users}.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{literature-review/images/butz-methodology}}
\caption{Overview of methodology followed by \citet{Butz2018}.}
\label{fig:butz-methodology}
\end{figure}

The document plan for the argumentation follows the same chain of strongest dependencies constructed in the \textit{knowledge base} until the expert disagrees; at that point, the user is presented with an alternative \enquote{MPE}.
An example of how the document plan may look after interaction with the user is shown in Figure \ref{fig:butz-tree}.
All the natural language phrasing is generated via boilerplates that take care of realising both the micro-planning phase and the generation of the text.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{literature-review/images/butz-tree}}
\caption{\textit{Document plan} generated from the \textit{probability tree} \citep{Butz2018}.}
\label{fig:butz-tree}
\end{figure}

The authors recognise that such chains of deduction could become long and cognitively overloading in the case of larger BNs, as every variable in the tree is explained by all its ancestors.
A solution they propose is that of \textit{pruning} the probability tree by excluding d-separated nodes and those under a certain threshold of significance.
They also adapt some methods from literature to perform \textit{conflict analysis} i.e., only variables that contribute positively to the explanation are maintained in the document plan.

On the whole, \citet{Butz2018} offer a compelling explanation method for BNs by building on an established tradition of enabling explainability through dialogue.
The work, though, takes some methodological missteps and also continues the \enquote{sin} of not validating its claims on real users, which is one of the primary gaps in the xAI field, as identified in the previous sections.

### 2019-11-13 literature-review/summary_lr.tex
Thesis complete

\section{Summary} \label{sec:literature-review-summary}
The findings outlined in this chapter refer to the concept of explainability of Machine Learning models, to its importance, to ways of evaluating it and to explainability in the specific case of Bayesian Networks.
The concept of explainability is central to the field of Explainable AI (xAI), whose goal is to make Machine Learning systems \enquote{interpretable}.
Explainability can briefly be defined as the property of a system that is able to \enquote{explain or to present in understandable terms to a human} \citep{dosilovic2018} its outputs.
Two main classes of explainable models have been identified by \citet{mittelstadt2019explaining}: ante-hoc or transparent and post-hoc interpretable ones; the formers are inherently inspectable in their inner workings while the latters are made understandable by way of extra techniques.
These two classes have also been refined by \citet{doshi2017towards} into a four-tier taxonomy consisting of: \textit{opaque}, \textit{interpretable}, \textit{comprehensible} and \textit{explainable systems}.
An opaque system, also know as a \enquote{black-box model}, is one whose inner workings are not inspectable from the outside; an interpretable system corresponds to an ante-hoc interpretable one; a comprehensible one emits extra information together with its output; an explainable system explicitly outputs a human-understandable line of reasoning aimed at clarifying its workings.

Explainability has become a central concept to the field of AI as a whole; as ML models take over more and more functions in our societies, the pressure for them to be able to explain their decisions is increased accordingly.
The \textit{General Data Protection Regulation} (GDPR) that became effective in 2018 was viewed by many as increasing the societal pressure to make systems explainable; many may have been mistaken as regards the actual rules mandated by the regulation - that aren't really prescribing a broad \enquote{right to an explanation} \citep{edwards2018enslaving} - but nonetheless the feeling of urgency is sure to increase the focus of both researchers and laypeople.
In general, explainability is framed as an issue of moral necessity as it easy to find a long series of situations where ML models displayed covert bias or what we would regard as bad moral judgement.

There are all manner of ways to measure explainability and these can be classified into a three-layer taxonomy \citep{doshi2017towards} based on the assumption that the best type of evaluation is the one that most involves humans.
The three classes are \textit{functionally-grounded evaluation}, \textit{human-grounded evaluation} and \textit{application-grounded evaluation}, ordered from the one least involving real humans to the one where the presence of the human-in-the-loop is greatest.
As the involvement of humans in evaluating models' explanations increases, so does the cost of such an experiment and its specificity, as the highest evaluation level necessarily entails the collaboration of domain-experts on specific tasks.
A parallel taxonomy identifies: methods to explain black box models, methods to explain black box outcomes, methods to inspect black boxes and methods to design transparent boxes.
An overarching notion that has been stressed is that \textit{explainability is a graded notion} that depends on the knowledge and expertise of the particular user: different users, with different expertises and backgrounds, may rate the quality of a same explanation very differently.

Bayesian Networks (BNs) have enjoyed widespread appeal in mission-critical domains like that of medicine and thus the drive to develop methods to explain their outputs has always been strong.
BNs have three main elements that necessitate an explanation \citep{lacave2002review}: the \textit{knowledge base}, the \textit{reasoning process} and the \textit{evidence propagated}.
Explaining the first \enquote{consists of determining which values of the unobserved variables justify the available evidence} and is done by solving the Most Probable Explanation (MPE) problem.
For the second, a static explanation of the BN is achieved by displaying it graphically or verbally.
The last element is explained by showing the reasoning that brought the BN to give the outputs it did that can be achieved by providing a justification for its outputs, for the results it did not give or via hypothetical reasoning.
The fact that BNs are able to naturally support counterfactual reasoning, combine single variables into composite outputs and model causality puts them at an advantage compared to other ML systems when generating an effective explanation for a user.
This is because the capabilities of a BN enable it to generate explanations that are uniquely suited to out psychological biases and expectations of what an explanation should entail.

Some of the main gaps that were found during the review of the literature relate to how there is still a great confusion in the field of xAI regarding what an explanation really is and thus what constitutes a good instance of it.
There is also a prevalent methodological confusion, as different authors use terms in incongruous ways, for example sometimes \textit{intepretation} is taken to mean \textit{explanation} while in other cases they refer to different concepts, for example in the taxonomy of interpretable systems proposed by \citet{doshi2017towards}.
This naturally makes it difficult for the field to converge onto methods to evaluate such explanations and this is reflected in the barrage of methods present in the literature, each one focused only on a particular system or instance of model.
This confusion is exacerbated by a seeming lack of interest or awareness of xAI researchers for the sizeable corpus of work in psychology, philosophy, social sciences, neuroscience and human computer interaction that has already investigated the nature of explanations, what desiderata they may possess and which are most effective.
In fact, the great majority of proposed approaches is only focused on proving formal explainability and neglects the human side that is naturally present in any explanation; there has been little work carried out to validate approaches in real settings with real domain experts so many explainability methods are substantiated only at theoretical level.
The underlying issue that has been seen to run transversely across the various concepts investigated in this chapter is best summarised by the idea that \enquote{inmates are running the asylum}, meaning that individual researchers are claiming that their models are interpretable referring only to their own personal views and biases and not to established literature and methods.
It would be hard for them to do otherwise, as the field of xAI seems, at present, to be a collection of diverging strands without a comprehensive program able to help it converge onto its stated goal: to make Machine Learning systems understandable by their users and thus increase their social utility and acceptance.

>>>>

\section{Summary} \label{sec:literature-review-summary}
The findings outlined in this chapter refer to the concept of explainability of machine learning models, to its importance, to ways of evaluating it and to explainability in the specific case of Bayesian networks.
The concept of explainability is central to the field of explainable AI, whose goal is to make machine learning systems \enquote{interpretable} or \enquote{explainable}.
\textit{Explainability} can briefly be defined as the property of a system that is able to \enquote{explain or to present in understandable terms to a human} \citep{dosilovic2018} its outputs.
Two main classes of explainable models have been identified by \citet{mittelstadt2019explaining}: \textit{ante-hoc or transparent} and \textit{post-hoc interpretable} ones; the former are inherently inspectable in their inner workings while the latter are made understandable by way of extra techniques.
These two classes have also been refined by \citet{doshi2017towards} into a four-tier taxonomy consisting of: \textit{opaque}, \textit{interpretable}, \textit{comprehensible} and \textit{explainable systems}.
An opaque system, also know as a \enquote{black-box model}, is one whose inner workings are not inspectable from the outside; an interpretable system corresponds to an ante-hoc interpretable one; a comprehensible one emits extra information together with its output; an explainable system explicitly outputs a human-understandable line of reasoning aimed at clarifying its workings.

Explainability has become a central concept to the field of AI as a whole; as ML models take over more and more functions in our societies, the pressure for them to be able to explain their decisions is increased accordingly.
The \textit{General Data Protection Regulation} that became effective in 2018 was viewed by many as increasing the societal pressure to make systems explainable; many may have been mistaken as regards the actual rules mandated by the regulation - that aren't really prescribing a broad \enquote{right to an explanation} \citep{edwards2018enslaving} - but nonetheless the feeling of urgency is sure to increase the focus of both researchers and laypeople.
In general, explainability is framed as an issue of moral necessity as it easy to find a long series of situations where ML models displayed covert bias or what we would regard as bad moral judgement.

There are all manner of ways to measure the quality of an explaination and these can be classified into a three-layer taxonomy \citep{doshi2017towards} based on the assumption that the best type of evaluation is the one that most involves humans.
The three classes are \textit{functionally-grounded evaluation}, \textit{human-grounded evaluation} and \textit{application-grounded evaluation}, ordered from the one least involving real humans to the one where the presence of the human-in-the-loop is greatest.
As the involvement of humans in evaluating models' explanations increases, so does the cost of such an experiment and its specificity, as the highest evaluation level necessarily entails the collaboration of domain-experts on specific tasks.
A parallel taxonomy identifies: methods to explain black box models, methods to explain black box outcomes, methods to inspect black boxes and methods to design transparent boxes.
An overarching notion that has been stressed is that \textit{explainability is a graded notion} that depends on the knowledge and expertise of the particular user: different users, with different expertises and backgrounds, may rate the quality of a same explanation very differently.

Bayesian networks (BNs) have enjoyed widespread appeal in mission-critical domains like that of medicine and thus the drive to develop methods to explain their outputs has always been strong.
BNs have three main elements that necessitate an explanation: the \textit{evidence}, the \textit{model} and the \textit{reasoning}. \citep{lacave2002review}
Explaining the first \enquote{consists of determining which values of the unobserved variables justify the available evidence} and is done by solving the most probable explanation problem.
For the second, a static explanation of the BN is achieved by displaying it graphically or linguistically.
The last element is explained by showing the reasoning that brought the BN to give the outputs it did and can be achieved by providing a justification for its outputs, for the results it did not give or via hypothetical reasoning.
The fact that BNs are able to naturally support counterfactual reasoning, combine single variables into composite outputs and model causality puts them at an advantage compared to other ML systems when generating an effective explanation for a user.
This is because the capabilities of a BN enable it to generate explanations that are uniquely suited to our psychological biases and expectations of what an explanation should entail.

Some of the main gaps that were found during the review of the literature relate to how there is still a great confusion in the field of xAI regarding what an explanation really is and thus what would constitute a good instance of it.
There is also a prevalent methodological confusion, as different authors use terms in incongruous ways, for example sometimes \textit{intepretation} is taken to mean \textit{explanation} while in other cases they refer to different concepts, for example in the taxonomy of interpretable systems proposed by \citet{doshi2017towards}.
This naturally makes it difficult for the field to converge onto methods to evaluate such explanations and this is reflected in the barrage of methods present in the literature, each one focused only on a particular system or model.
This confusion is exacerbated by a seeming lack of interest or awareness of xAI researchers for the sizeable corpus of work in psychology, philosophy, social sciences, neuroscience and human computer interaction that has already investigated the nature of explanations, what desiderata they may possess and which are most effective.
In fact, the great majority of proposed approaches is only focused on proving formal explainability and neglects the human side that is naturally present in any explanation; there has been little work carried out to validate approaches in real settings with real domain experts so many explainability methods are substantiated only at theoretical level.
The underlying issue that has been seen running transversely across the various concepts investigated in this chapter is best summarised by the idea that \enquote{inmates are running the asylum}, meaning that individual researchers are claiming that their models are interpretable referring only to their own personal views and biases and not to established literature and methods.
It would be hard for them to do otherwise, as the field of xAI seems, at present, to be a collection of diverging strands without a comprehensive program able to help it converge onto its stated goal: to make machine learning systems understandable by their users and thus increase their social utility and acceptance.
